---
title: "Scraping CDC"
author: "Hannah & Bennet"
date: "5/6/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
library(rvest)
```


## Sources (not data we read in, just information and statistics we collected for conclusion)

https://www.cdc.gov/heartdisease/facts.htm
Heart disease fact page, US.

https://www.cdc.gov/tobacco/data_statistics/fact_sheets/fast_facts/index.htm
Smoking numbers, US.

https://www.cdc.gov/obesity/data/prevalence-maps.html
Obesity displayed on US map, used to observe patterns

https://www.ncdemography.org/2014/06/16/mortality-and-cause-of-death-1900-v-2010/
1900 vs 2010 mortality numbers, US.

https://jamanetwork.com/journals/jama/fullarticle/2778234
Death numbers, 2015-2020, US.

https://www.safetynewsalert.com/number-of-accidental-deaths-hits-new-high/
What are accidental death causes?

https://www.cdc.gov/nchs/fastats/obesity-overweight.htm
Obesity and overweight numbers, US.

https://www.who.int/news-room/fact-sheets/detail/the-top-10-causes-of-death
Top causes of death globally, breakdown by income level.

## CDC Total Deaths


```{r}
#creating a df from CDC chart of top 10 causes of death in U.S.

CDC <- read_html("https://www.cdc.gov/nchs/fastats/leading-causes-of-death.htm")
CDC_table <- data.frame(cause = c("Heart disease", "Cancer", "Accidents", "Chronic lower respiratory diseases", "Stroke", "Alzheimer's disease", "Diabetes", "Nephritis, nephrotic syndrom, and nephrosis", "Influenza and pneumonia", "Suicide"),
                 deaths = c("659,041", "599,601", "173,040", "156,979", "150,005", "121,499", "87,647", "51,565", "49,783", "47,511"))

#creating csv from top 10 causes of death data
write_csv(CDC_table,"~/Mscs 264b S21/Project/Hannah and Bennet/CDC_table.csv")


```

```{r}
#reading csv, converting data into tibble
CDC_chart <- read_csv("~/Mscs 264b S21/Project/Hannah and Bennet/CDC_table.csv")

#initial test graph
CDC_chart %>%
  ggplot() +
  geom_point(mapping = aes(x = cause, y = deaths))
```

